### 2018-05-28 write-up/evaluation.tex
Remove onehalfspacing
Fix many small writing errors

\prechapter{%
    I will evaluate the central premise of this thesis: linear types are a
    practical and usable tool to help programmers write readable (equational
    and algebraic), safe (with respect to aliasing, read/write permissions,
    memory re-use and deallocation) and explicit (with respect to memory
    allocation) code that (1) is more memory-efficient than code written using
    high-level linear algebra libraries and (2) performs just as predictably as
    code written using low-level linear-algebra libraries. I will also
    elaborate on the qualitative benefits of using linear types to write
    linear-algebra programs.
}%

@@

To test whether code written using LT4LA performs just as predictably as code
written using low-level linear-algebra libraries, I wrote two implementations
of a Kalman filter:

@@

I measured execution time, in micro-seconds, against an exponentially (powers
of 5) increasing scaling factor for matrix size parameters $n=5$ and $k=3$.
For small scaling factors (1, 5, 25), I used the Core\_bench micro-benchmarking
library, for larger factors (125 and greater), I used the \texttt{getrusage}
system call (called \ltfla{Unix.times} in OCaml), sandwiched between calls to
\ltfla{Gc.full_major} to ensure no garbage collection took place during the
measurements.  Core\_bench performs a linear-regression (here, time against
batch-size) so includes 95\% confidence-interval with $R^2$. Larger scales have
errors reported to $\pm \sigma$ (one standard-deviation).

@@

Inspecting the trace for LT4LA shows 6 calls to `empty' (function for
allocating new matrices): 4 temporaries plus 2 for storing the results. From
the CBLAS implementation (Figure~\ref{fig:cblas_kalman}), we can see that it
too has 6 calls to malloc: 4 temporaries plus 2 to storing the results.

@@

For $n=5$, CBLAS was faster ($24\mu s$, 526 samples) than LT4LA ($41 \mu s$,
466 samples).  For $n=25$, and around 350 samples each, CBLAS was again faster
($104 \mu s$) than LT4LA ($133 \mu s$). The 95\% confidence-intervals for these
measurements differ from the means by at most $2 \mu s$. For $n=125$, and
around 110 samples, CBLAS was \emph{slower} ($1803 \mu s\, [1746,\, 1867]$)
than LT4LA ($1678 \mu s\, [1646,\, 1714]$). For $n=625$ and 1000 samples each,
they took a \emph{very similar} amount of time: LT4LA took $180.5 \pm 38 \,ms$
and CBLAS took $188 \pm 36 \, ms$. Despite the large sample size, the
standard-deviation is still quite high; however, \emph{because} of the large
sample size, the $p$-value (Welch's t-test) is very small ($p < .05$),
suggesting that the difference in the means is statistically significant.
Lastly, for $n=3125$ and 15 samples each, LT4LA and CBLAS continued to take a
similar amount of time ($16.1 \pm 0.19 \,s$ and $15.7 \pm 0.53 \,s$
respectively, $p<.05$ using Welch's t-test).

@@

For small matrix sizes, LT4LA and CBLAS execution times differ. I suspect this
is due to large sample sizes causing more allocations and thus more
garbage-collector unpredictability (which the linear-regression \emph{did not}
take into account; multi-variate regressions are not fully supported by
Core\_bench yet). As matrix size increases, execution times of LT4LA become
very similar to those of CBLAS.

Therefore, by using LT4LA, it is possible to write linear-algebra programs that
perform just as predictably (with regards to memory usage and execution time)
as code written using low-level linear algebra libraries, but gain readability
(equational and algebraic expressions) and safety (with respect to aliasing
and read/write permissions).

@@

To test whether code written using LT4LA is more memory efficient than code
written using high-level linear-algebra libraries, I wrote two\footnote{I
attempted a third implementation using Owl's \emph{Lazy} module but it did
not support matrix inversion at the time of writing.} more implementations
of a Kalman filter:

@@

let owl_kalman ~sigma ~h ~mu ~r ~data =
  let open Owl.Mat in
  let ( * ) = dot in
  let h' = transpose h in
  let sigma_h' = sigma * h' in
  let x = sigma_h' * (inv @@ r + h * sigma_h') in
  let new_mu = mu + x * (h * mu - data) in
  let new_sigma = sigma - x * h * sigma in
  ((sigma, (h, (mu, (r, data)))), (new_mu, new_sigma))
;;
    \end{minted}
    \caption{Implementations of a Kalman filter using Owl, top one using a
        Cholesky decomposition, bottom one using idiomatic Owl. Owl does not
        yet provide a non-mutating `potrs' function, so I wrote my own which
        returns a mutated copy of its argument instead.}\label{fig:chol_owl_kalman}

@@

For LT4LA and CBLAS the total amount of memory allocated for temporaries is $n
+ n^2 + nk + k^2$. The difference between these two implementations and the
idiomatic Owl implementation is $k(3n+2k+2)$ words. Hence, by using LT4LA, it
is possible to have the readability (equational and algebraic expressions) and
safety of high-level linear algebra libraries, whilst gaining precise and
explicit control over memory allocation and re-use.

@@

I chose the example of a Kalman filter because it is used in the real world,
consists purely of a sequence of matrix expressions and produces many
unnecessary temporary matrices when implemented idiomatically in a high-level
linear-algebra library. It is good for isolating the key differences between
not having and having linear types to help a programmer safely manage memory
and aliasing, whilst excluding other aspects also important to real world
linear-algebra programs such as control flow or blocking.

@@

A graph of the execution times (with error bars which are present but quite
small) \emph{all four} implementations is show in Figure~\ref{fig:timings_all}.

@@

@@

A \textbf{positive take-away} from all of this is that LT4LA at least gives a
programmer \emph{choice} and \emph{control} about how to optimise their program
to their needs because (1) it met its goal of enabling a programmer to write
code that performs just as predictably as code written using lower-level
linear-algebra libraries (2) it does so \emph{without} the associated dangers
of unenforced and uncheckable rules around aliasing, read/write permissions,
memory allocation, re-use and deallocation that would come with using C,
Fortran, or the unsafe primitives of Owl.

@@

We have already seen (last chapter and this chapter) that linear types help
programmers write readable, safe (with respect to aliasing, read/write
permissions, memory re-use and deallocation) and explicit (with respect to
memory allocation) code. To justify their why I think they are a
\emph{practical} and \emph{usable} way to do so, I will elaborate on some of
the qualitative benefits I experienced whilst using LT4LA.

Prior to this project, I had no experience with linear-algebra libraries or the
problem of matrix expression compilation. As such, I based my initial LT4LA
implementation of a Kalman filter using BLAS and LAPACK, on a popular GitHub
gist of a Fortran implementation, one that was automatically generated from
SymPy's matrix expression compiler~\cite{rocklin_thesis}.

Once I translated the implementation, I attempted to compile it and found that
(to my surprise) it did not type-check. This was because the original
implementation contained incorrect aliasing, did not adhere to Fortran's
read/write permissions (with respect to \texttt{intent} annotations
\texttt{in}, \texttt{out} and \texttt{inout}) and unused and unnecessary
temporaries, all of which were now highlighted by LT4LA's type system.

The original implementation used 6 temporaries, one of which was immediately
spotted as not being used due to linearity. It also contained two variables
which were marked as \texttt{intent(in)} but would have been written over by
calls to `gemm', spotted by the fractional-capabilities feature. Furthermore,
it used a matrix \emph{twice} in a call to `symm', once with a read permission
but once with a \emph{write} permission.  Fortran assumes that any parameter
being written to is not aliased and so this call was not only incorrect, but
illegal according to the standard, both aspects of which were captured by
linearity and fractional-capabilities. Lastly, it contained another unnecessary
temporary, however one that was not obvious without linear types. To spot it, I
first performed live-range splitting (checked by linearity) by hoisting calls
to \ltfla{freeM} and then annotated the freed matrices with their dimensions.
After doing so and spotting two disjoint live-ranges of the same size, I
replaced a call to \ltfla{freeM} followed by allocating call to \ltfla{copy}
with one, in-place call to \ltfla{copyM_to}. I believe the ability to boldly
refactor code which manages memory is good evidence of the usefulness of
linearity as a tool for programming.

@@

Writing a linear-algebra program using LT4LA combines the best of low- and
high-level linear-algebra libraries: it gives a programmer \emph{explicit
control} (over read/write permissions, aliasing and memory allocation, re-use
and deallocation), readability (equational and algebraic expressions)
\emph{and} safety (automatic checking). Programs written using it \emph{signal
precise intent} to the reader and compiler and perform similarly to equivalent
programs written directly using lower-level linear-algebra libraries. In turn,
such programs can be \emph{checked automatically} against their intent,
especially when \emph{refactoring} or \emph{rewriting} code. Although any
expert \emph{could} have followed the same line of reasoning laid out above,
and arrived at the same program, LT4LA's type system enables a non-expert
(yours truly) to do the same with increased confidence in the
result\footnote{This does not preclude testing code by actually
\emph{executing} it, but definitely \emph{complements} it.} by checking said
reasoning.

>>>>

\prechapter{%
    I will evaluate the central premise of this thesis: linear types are a
    practical and usable tool to help programmers write readable (equational
    and algebraic), safe (with respect to aliasing, read/write permissions,
    memory re-use and deallocation) and explicit (with respect to memory
    allocation) code that (1) is more memory-efficient than code written using
    high-level linear algebra libraries and (2) performs just as predictably as
    code written using low-level linear algebra libraries. I will also
    elaborate on the qualitative benefits of using linear types to write
    linear algebra programs.
}%

@@

To test whether code written using LT4LA performs just as predictably as code
written using low-level linear algebra libraries, I wrote two implementations
of a Kalman filter:

@@

I measured execution time, in micro-seconds, against an exponentially (powers
of 5) increasing scaling factor for matrix size parameters $n=5$ and $k=3$.
For small scaling factors (1, 5, 25), I used the Core\_bench micro-benchmarking
library, for larger factors (125 and greater), I used the \texttt{getrusage}
system call (called \ltfla{Unix.times} in OCaml), sandwiched between calls to
\ltfla{Gc.full_major} to minimise the effects of garbage-collection.
Core\_bench performs a linear regression (here, time against batch-size) so
includes a 95\% confidence-interval with $R^2$. Larger scales have
errors reported to $\pm \sigma$ (one standard-deviation).

Inspecting the trace for LT4LA shows 6 calls to `empty' (function for
allocating new matrices): 4 temporaries plus 2 for storing the results. From
the CBLAS implementation (Figure~\ref{fig:cblas_kalman}), we can see that it
too has 6 calls to malloc: 4 temporaries plus 2 for storing the results.

@@

@@

For $n=5$, CBLAS was faster ($24\mu s$, 526 samples) than LT4LA ($41 \mu s$,
466 samples).  For $n=25$, and around 350 samples each, CBLAS was again faster
($104 \mu s$) than LT4LA ($133 \mu s$). The 95\% confidence-intervals for these
measurements differ from the means by at most $2 \mu s$. For $n=125$, and
around 110 samples each, CBLAS was \emph{slower} ($1803 \mu s\, [1746,\, 1867]$)
than LT4LA ($1678 \mu s\, [1646,\, 1714]$). For $n=625$ and 1000 samples each,
they took a \emph{very similar} amount of time: LT4LA took $180.5 \pm 38 \,ms$
and CBLAS took $188 \pm 36 \, ms$. Despite the large sample size, the
standard-deviation is still quite high; however, \emph{because} of the large
sample size, the $p$-value (Welch's t-test) is very small ($p < .05$),
suggesting that the difference in the means is statistically significant.
Lastly, for $n=3125$ and 15 samples each, LT4LA and CBLAS continued to take a
similar amount of time ($16.1 \pm 0.19 \,s$ and $15.7 \pm 0.53 \,s$
respectively, $p<.05$ using Welch's t-test).

@@

For small matrix sizes, LT4LA and CBLAS execution times differ. I suspect this
is due to large sample sizes causing more allocations and thus more
garbage-collector unpredictability (which the linear regression \emph{did not}
take into account; multi-variate regressions are not fully supported by
Core\_bench yet). As matrix size increases, execution times of LT4LA become
very similar to those of CBLAS.

Therefore, by using LT4LA, it is possible to write linear algebra programs that
perform just as predictably (with regards to memory usage and execution time)
as code written using low-level libraries, but gain readability (equational and
algebraic expressions) and safety (with respect to aliasing
and read/write permissions).

@@

To test whether code written using LT4LA is more memory-efficient than code
written using high-level libraries, I wrote two\footnote{I attempted a third
implementation using Owl's \emph{Lazy} module but it did not support matrix
inversion at the time of writing.} more implementations of a Kalman filter:

@@

let owl_kalman ~sigma ~h ~mu ~r ~data =
  let open Owl.Mat in
  let ( * ) = dot in
  let h' = transpose h in
  let sigma_h' = sigma * h' in
  let x = sigma_h' * (inv @@ r + h * sigma_h') in
  let new_mu = mu + x * (h * mu - data) in
  let new_sigma = sigma - x * h * sigma in
  ((sigma, (h, (mu, (r, data)))), (new_mu, new_sigma))
;;
    \end{minted}
    \caption{Implementations of a Kalman filter using Owl, top one using a
        Cholesky decomposition, bottom one using idiomatic Owl. Owl does not
        yet provide a non-mutating `potrs' function, so I wrote my own which
        returns a mutated \emph{copy} of its argument
        instead.}\label{fig:chol_owl_kalman}

@@

For LT4LA and CBLAS the total amount of memory allocated for temporaries is $n
+ n^2 + nk + k^2$. The difference between these two implementations and the
idiomatic Owl implementation is $k(3n+2k+2)$ words. Hence, by using LT4LA, it
is possible to have the readability (equational and algebraic expressions) and
safety of high-level libraries, and gain precise and explicit control over
memory allocation and re-use.

@@

I chose the example of a Kalman filter because it is used in the real world,
consists purely of a sequence of matrix expressions and produces many
unnecessary temporary matrices when implemented idiomatically in a high-level
library. It is good for isolating the key differences between not having and
having linear types to help a programmer safely manage memory, aliasing, and
read/write permissions, whilst excluding other aspects also important to real
world linear algebra programs such as control flow or blocking.

@@

A graph of the execution times (with error bars which are present but quite
small) of \emph{all four} implementations is show in Figure~\ref{fig:timings_all}.

@@

    \item CBLAS/LT4LA use Cholesky decomposition: it performs \emph{half} as
        many operations as LU decomposition (as used by Owl) under the
        assumption that its argument is a symmetric matrix.

@@

A \textbf{positive take-away} from all of this is that LT4LA at least gives a
programmer \emph{choice} and \emph{control} about how to optimise their program
to their needs because (1) it met its goal of enabling a programmer to write
code that performs just as predictably as code written using low-level
libraries (2) it does so \emph{without} the associated dangers of unenforced
and uncheckable rules regarding aliasing, read/write permissions, memory
allocation, re-use and deallocation that would come with using C, Fortran, or
the unsafe primitives of Owl.

@@

We have already seen (last chapter and this chapter) that linear types help
programmers write readable, safe (with respect to aliasing, read/write
permissions, memory re-use and deallocation) and explicit (with respect to
memory allocation) code. To justify their why I think they are a
\emph{practical} and \emph{usable} way to do so, I will elaborate on
the qualitative benefits I experienced whilst using LT4LA.

Prior to this project, I had no experience with linear algebra libraries or the
problem of matrix expression compilation. As such, I based my initial LT4LA
implementation of a Kalman filter using BLAS and LAPACK, on a popular GitHub
gist of a Fortran implementation, one that was automatically generated from
SymPy's matrix expression compiler~\cite{rocklin_thesis}.

Once I translated the implementation from Fortran to LT4LA, I attempted to
compile it and found that (to my surprise) it did not type-check. This was
because the original implementation contained incorrect aliasing, unused and
unnecessary temporaries, and did not adhere to Fortran's read/write permissions
(with respect to \texttt{intent} annotations \texttt{in}, \texttt{out} and
\texttt{inout}) all of which were now highlighted by LT4LA's type system.

The original implementation used 6 temporaries, one of which was immediately
spotted as never being used due to linearity. It also contained two variables
which were marked as \texttt{intent(in)} but would have been written over by
calls to `gemm', spotted by the fractional-capabilities feature. Furthermore,
it used a matrix \emph{twice} in a call to `symm', once with a read permission
but once with a \emph{write} permission.  Fortran assumes that any parameter
being written to is not aliased and so this call was not only incorrect, but
illegal according to the standard, both aspects of which were captured by
linearity and fractional-capabilities. Lastly, it contained another unnecessary
temporary, however one that was not obvious without linear types. To spot it, I
first performed live-range splitting (checked by linearity) by hoisting calls
to \ltfla{freeM} and then annotated the freed matrices with their dimensions.
After doing so and spotting two disjoint live-ranges of the same size, I
replaced a call to \ltfla{freeM} followed by allocating call to \ltfla{copy}
with one, in-place call to \ltfla{copyM_to}. I believe the ability to boldly
refactor code which manages memory is good evidence of the usefulness of
linearity as a tool for programming.

@@

Writing a linear algebra program using LT4LA combines the best of low- and
high-level libraries: it gives a programmer \emph{explicit
control} (over read/write permissions, aliasing and memory allocation, re-use
and deallocation), readability (equational and algebraic expressions)
\emph{and} safety (automatic checking). Programs written using it \emph{signal
precise intent} to the reader and compiler and perform similarly to equivalent
programs written directly using low-level libraries. In turn,
such programs can be \emph{checked automatically} against their intent,
especially when \emph{refactoring} or \emph{rewriting} code. Although any
expert \emph{could} have followed the same line of reasoning laid out above,
and arrived at the same program, LT4LA's type system enables a non-expert
(yours truly) to do the same with increased confidence in the
result\footnote{This does not preclude testing code by actually
\emph{executing} it, but definitely \emph{complements} it.} by checking said
reasoning.
